---
title: "NDVI Data"
format: pdf
---

### libraries and setup 
```{r}
library(rstac)
library(gdalcubes)
library(stars)
library(tmap)
library(dplyr)
library(stringr)
library(sf)
gdalcubes::gdalcubes_options(parallel = TRUE)
```

## Pre-Fire Data
### defining our area of interest in our bounding box, and get list of metadata
```{r}
box <- c(
  xmin = -122.3,
  ymin = 37.90,
  xmax = -122.2,
  ymax = 38.0
)

#box <- c(xmin= -122.24668, ymin= 37.90971, xmax= -122.24689, ymax=37.90984)

start_date <- "2023-06-01"
end_date <- "2023-09-01"
items <-
  stac("https://earth-search.aws.element84.com/v1") |>
  stac_search(
    collections = "sentinel-2-l2a",
    bbox = box,
    datetime = paste0(start_date, "T00:00:00Z/", end_date, "T23:59:59Z"),
    limit = 100
  ) |>
  ext_query("eo:cloud_cover" < 20) |>
  post_request()

```


### passing to rstac to extract data for our bounding box
```{r}
col <- stac_image_collection(
  items$features,
  asset_names = c("red", "nir", "scl")
)

#srs ="EPSG:32610",
cube <- cube_view(srs ="EPSG:4326",
                  extent = list(t0 = start_date, t1 = end_date,
                                left = box[1], right = box[3],
                                top = box[4], bottom = box[2]),
                  dx = 0.0001, dy = 0.0001, dt = "P1D",
                  aggregation = "median", resampling = "average")


mask <- image_mask("scl", values=c(3, 8, 9)) # mask clouds and cloud shadows


data <- raster_cube(col, cube, mask = mask)

```

### calcuating NDVI
```{r}
ndvi <- data |>
  select_bands(c("red", "nir")) |>
  apply_pixel("(nir-red)/(nir+red)", "NDVI") |>
  reduce_time(c("mean(NDVI)"))

ndvi_stars <- st_as_stars(ndvi)
```

### checks/ summary stats to ensure NDVI calculated properly
```{r}
# extract vector of values and compute summary stats (no summary() print)
#vals <- as.data.frame(ndvi_stars, xy = FALSE)[["NDVI_mean"]]
#stats <- c(
#  min = min(vals, na.rm = TRUE),
#  p25 = quantile(vals, 0.25, na.rm = TRUE),
#  median = median(vals, na.rm = TRUE),
#  mean = mean(vals, na.rm = TRUE),
#  p75 = quantile(vals, 0.75, na.rm = TRUE),
#  max = max(vals, na.rm = TRUE),
#  prop_na = mean(is.na(vals))
#)
#stats

# inspect band names in the raw raster cubes (may be large; this only lists names)
#names(st_as_stars(data))[1:10]        # names present in pre-fire cube --> result: "red", "nir", "NA", "NA"   may need to run other checks for why mask scl not present

```

### plotting the result
```{r}
mako <- tm_scale_continuous(values = viridisLite::mako(30))
fill <- tm_scale_continuous(values = "Greens")

tm_shape(ndvi_stars) + tm_raster(col.scale = mako)
```

### Post-fire data

```{r}
# Read your CSV with plot coordinates
plots <- read.csv("latlong.csv")
# Convert to sf object
plots_sf <- st_as_sf(plots, coords = c("long", "lat"), crs = 4326)

# Extract NDVI values
calculate_ndvi <- st_extract(ndvi_stars, plots_sf, FUN = mean)
ndvi_values <- calculate_ndvi$NDVI_mean

# Combine with plot IDs
plots_ndvi <- cbind(plots, pre_fire_NDVI = ndvi_values)
plots_ndvi

```

## Post-Fire
```{r}
start_post_fire <- "2024-08-20"
end_post_fire <- "2024-10-20"
items_post <-
  stac("https://earth-search.aws.element84.com/v1") |>
  stac_search(
    collections = "sentinel-2-l2a",
    bbox = box,
    datetime = paste0(start_post_fire, "T00:00:00Z/", end_post_fire, "T23:59:59Z"),
    limit = 100
  ) |>
  ext_query("eo:cloud_cover" < 20) |>
  post_request()

```

```{r}
col_post <- stac_image_collection(
  items_post$features,
  asset_names = c("red", "nir", "scl")
)

cube_post <- cube_view(srs ="EPSG:4326",
                  extent = list(t0 = start_post_fire, t1 = end_post_fire,
                                left = box[1], right = box[3],
                                top = box[4], bottom = box[2]),
                  dx = 0.0001, dy = 0.0001, dt = "P1D",
                  aggregation = "median", resampling = "average")


mask <- image_mask("scl", values=c(3, 8, 9)) # mask clouds and cloud shadows


data_post <- raster_cube(col_post, cube_post, mask = mask)

```

```{r}
ndvi_post <- data_post |>
  select_bands(c("red", "nir")) |>
  apply_pixel("(nir-red)/(nir+red)", "NDVI") |>
  reduce_time(c("mean(NDVI)"))

ndvi_stars_post <- st_as_stars(ndvi_post)
```

### plotting the result
```{r}
mako_post <- tm_scale_continuous(values = viridisLite::mako(30))
fill_post <- tm_scale_continuous(values = "Greens")

tm_shape(ndvi_stars_post) + tm_raster(col.scale = mako_post)
```

### now for our plots

```{r}

# Extract NDVI values
calculate_ndvi_post <- st_extract(ndvi_stars_post, plots_sf, FUN = mean)
ndvi_values_post <- calculate_ndvi_post$NDVI_mean

# Combine with plot IDs
all_ndvi <- cbind(plots_ndvi, post_fire_NDVI = ndvi_values_post)

#export to csv
write.csv(all_ndvi, "soil_plots_ndvi.csv", row.names = FALSE)
```

## ANOVA pre- vs post-fire
```{r}
treatment <- c("burn_heat", "burn_control", "burn_heat", "burn_control", "burn_heat", "burn_control", "burn_heat", "burn_control", "nonburn", "nonburn", "nonburn", "nonburn")

ndvi_data <- cbind(all_ndvi, treatment = treatment) |>
    relocate(treatment)

ndvi_data$NDVI_change <- ndvi_data$post_fire_NDVI - ndvi_data$pre_fire_NDVI
anova_model <- aov(NDVI_change ~ treatment, data = ndvi_data)
summary(anova_model)

```

### checking assumptions

``` {r}

plot(anova_model, which = 1) # Residuals vs Fitted
plot(anova_model, which = 2) # Normal Q-Q
kruskal.test(NDVI_change ~ treatment, data = ndvi_data)

```


## Trying with different CRS for better zoom

```{r}
# 1. Extract EPSG from Sentinel metadata
epsg_code <- items$features[[1]]$properties$`proj:epsg`

# 2. Transform bounding box from WGS84 → UTM
aoi <- st_as_sf(
  st_sfc(st_polygon(list(rbind(
    c(box["xmin"], box["ymin"]),
    c(box["xmax"], box["ymin"]),
    c(box["xmax"], box["ymax"]),
    c(box["xmin"], box["ymax"]),
    c(box["xmin"], box["ymin"])
  ))), crs = 4326
))
aoi_utm <- st_transform(aoi, paste0("EPSG:", epsg_code))
utm_box <- st_bbox(aoi_utm)

# 3. Build cube in UTM
cube_utm <- cube_view(
  srs = paste0("EPSG:", epsg_code),
  extent = list(
    t0 = start_date,
    t1 = end_date,
    left = utm_box["xmin"], right = utm_box["xmax"],
    top = utm_box["ymax"], bottom = utm_box["ymin"]
  ),
  dx = 20, dy = 20,     # ~20m grid
  dt = "P1D",
  aggregation = "median"
)

mask <- image_mask("scl", values=c(3, 8, 9)) # mask clouds and cloud shadows


utm_data <- raster_cube(col, cube_utm, mask = mask)

```

### calcuating NDVI
```{r}
utm_ndvi <- utm_data |>
  select_bands(c("red", "nir")) |>
  apply_pixel("(nir-red)/(nir+red)", "NDVI") |>
  reduce_time(c("mean(NDVI)"))

ndvi_stars_utm <- st_as_stars(utm_ndvi)

```

### plotting the result
```{r}

plots <- read.csv("latlong.csv")
library(sf)
plots_sf <- st_as_sf(plots, coords = c("long", "lat"), crs = 4326)
plots_utm <- st_transform(plots_sf, paste0("EPSG:", epsg_code))
ndvi_values <- st_extract(ndvi_stars_utm, plots_utm, FUN = mean)
plots_ndvi <- cbind(plots, NDVI = ndvi_values$NDVI_mean)



# Convert raster to data.frame for ggplot
ndvi_df <- as.data.frame(ndvi_stars_utm, xy = TRUE)

library(ggplot2)

# Calculate limits around plots (with a margin)
margin <- 50  # in UTM meters
xlim <- range(st_coordinates(plots_utm)[,1]) + c(-margin, margin)
ylim <- range(st_coordinates(plots_utm)[,2]) + c(-margin, margin)

ggplot() +
  geom_raster(data = ndvi_df, aes(x = x, y = y, fill = NDVI_mean)) +
  geom_sf(data = plots_utm, color = "red", size = 2) +
  coord_sf(xlim = xlim, ylim = ylim) +
  scale_fill_viridis_c(option = "mako") +
  labs(title = "Zoomed NDVI near Soil Plots") +
  theme_minimal()

```

```{r}
start_post_fire <- "2024-08-20"
end_post_fire <- "2024-10-20"
items_post <-
  stac("https://earth-search.aws.element84.com/v1") |>
  stac_search(
    collections = "sentinel-2-l2a",
    bbox = box,
    datetime = paste0(start_post_fire, "T00:00:00Z/", end_post_fire, "T23:59:59Z"),
    limit = 100
  ) |>
  ext_query("eo:cloud_cover" < 20) |>
  post_request()

```

```{r}
col_post <- stac_image_collection(
  items_post$features,
  asset_names = c("red", "nir", "scl")
)

# 2. Transform bounding box from WGS84 → UTM
aoi_ <- st_as_sf(
  st_sfc(st_polygon(list(rbind(
    c(box["xmin"], box["ymin"]),
    c(box["xmax"], box["ymin"]),
    c(box["xmax"], box["ymax"]),
    c(box["xmin"], box["ymax"]),
    c(box["xmin"], box["ymin"])
  ))), crs = 4326
))
aoi_utm_post <- st_transform(aoi, paste0("EPSG:", epsg_code))
utm_box_post <- st_bbox(aoi_utm)

# 3. Build cube in UTM
cube_utm_post <- cube_view(
  srs = paste0("EPSG:", epsg_code),
  extent = list(
    t0 = start_post_fire,
    t1 = end_post_fire,
    left = utm_box_post["xmin"], right = utm_box_post["xmax"],
    top = utm_box_post["ymax"], bottom = utm_box_post["ymin"]
  ),
  dx = 20, dy = 20,     # ~20m grid
  dt = "P1D",
  aggregation = "median"
)

utm_data_post <- raster_cube(col_post, cube_utm_post, mask = mask)


```
