---
title: "SVtilden_Code_emmaedit"
format: pdf
---

# Required packages to run script. If not on computer download and install.
``` {r}
library(ggplot2)
library(lubridate)
library(reshape2)
library(data.table)
library(dplyr)
library(gridExtra)
library(zoo)
library(plotly)
library(gridExtra)
library(data.table)
library(slider)
library(tidyverse)
```



# Code from Campbell sci to keep header, remove NANs, convert timestamp to correct format, and add column for file name as ID, 
``` {r}

importCSdata <- function(filename,RetOpt="data"){
  if(RetOpt=="info"){
    # bring in entire header of CSI TOA5 data file for metadata
    stn.info <- scan(file=filename,nlines=4,what=character(),sep="\r")
    return(stn.info)
  } else {
    # second line of header contains variable names
    header <- scan(file=filename,skip=1,nlines=1,what=character(),sep=",")
    # bring in data
    stn.data <- read.table(file=filename,skip=4,header=FALSE, na.strings=c("NAN"),sep=",")
    names(stn.data) <- header
    # add column of R-formatted date/timestamps
    stn.data$TIMESTAMP <- as.POSIXct(strptime(stn.data$TIMESTAMP,format="%m/%d/%Y %H:%M", tz="us/pacific"))
    stn.data$ID<-basename(filename)
    return(stn.data)}}

#set working directory
setwd("C:/Users/emmam/Downloads/senior thesis/DataForWorking")

# import all .dat files within folder
filenames=list.files(path=getwd(),pattern="[.]csv", recursive=T)

# bring in the files in a list using the Campbell function
list <- lapply(filenames, importCSdata)

#Combine all files in the list into a data frame. Select fill=T to merge files with different column numbers
data <-rbindlist(list, fill = TRUE)

#convert to data table
data <- data.table(data)

# Remove duplicates from data.
data <- data[!duplicated(data),]

#Order file based on timestamp
data<-data[order(TIMESTAMP),]


##SoilVUE
#Select columns for SoilVUE
SVburnt<-cbind(data[,c(1,69:356)])

#This will make it so there's only ONE column for temp, as opposed to various columns (goes from wide to long table format). Keep information that is pertinent to identify temp like timestamp and PP
sv <- SVburnt %>%
  melt(id.vars=c( "TIMESTAMP"), value.name="num")

sv=data.table(sv)
```

# bringing in the new data up to january and formatting it to match old
```{r}
#new data from jan

raw <- read_csv("C:/Users/emmam/Downloads/senior thesis/20250922_20260114.csv")

# Remove duplicates from data.
new <- raw[!duplicated(raw),]


##SoilVUE
#Select columns for SoilVUE
SVnew<-cbind(new[,c(1,69:356)])

# Convert new data to data.table format
jan_sv <- as.data.table(SVnew)

# Melt using data.table syntax (wide to long)
jan <- data.table::melt(jan_sv, 
                        id.vars = "TIMESTAMP", 
                        value.name = "num")

# --- FIX: Parse jan's TIMESTAMP to POSIXct before rbindlist ---
jan$TIMESTAMP <- as.POSIXct(strptime(jan$TIMESTAMP, format = "%m/%d/%Y %H:%M", tz = "US/Pacific"))                        

# Ensure both sv and jan have the same column names and types
jan$variable <- as.character(jan$variable)
sv$variable <- as.character(sv$variable)
```

# joining data set to be continuous from nov 2024 to jan 2026
``` {r}
# Bind the two tables together
sv_combined <- rbindlist(list(sv, jan), fill = TRUE)

# Parse variable names (same as before)
sv_combined[, c("rest", "plot", "depth", "avg") := tstrsplit(variable, "_", fixed = TRUE)]
sv_combined$rest <- NULL

# Assign PP, trt, measurement (same logic as before)
sv_combined[, PP := as.character(NA)]
sv_combined[, PP := fifelse(plot == "BH1" | plot == "BC1", "1", PP)]
sv_combined[, PP := fifelse(plot == "BH2" | plot == "BC2", "2", PP)]
sv_combined[, PP := fifelse(plot == "BH3" | plot == "BC3", "3", PP)]
sv_combined[, PP := fifelse(plot == "BH4" | plot == "BC4", "4", PP)]
sv_combined[, PP := fifelse(plot == "NB1", "5", PP)]
sv_combined[, PP := fifelse(plot == "NB2", "6", PP)]
sv_combined[, PP := fifelse(plot == "NB3", "7", PP)]
sv_combined[, PP := fifelse(plot == "NB4", "8", PP)]

sv_combined$trt <- NA
sv_combined[, trt := ifelse(plot == "BH1" | plot == "BH2" | plot == "BH3" | plot == "BH4", "H", trt)]
sv_combined[, trt := ifelse(plot == "BC1" | plot == "BC2" | plot == "BC3" | plot == "BC4", "C", trt)]
sv_combined[, trt := ifelse(plot == "NB1" | plot == "NB2" | plot == "NB3" | plot == "NB4", "H", trt)]

sv_combined$measurement <- NA
sv_combined[, measurement := ifelse(avg == "Avg(1)", "moisture", measurement)]
sv_combined[, measurement := ifelse(avg == "Avg(2)", "permitivity", measurement)]
sv_combined[, measurement := ifelse(avg == "Avg(3)", "temp", measurement)]
sv_combined[, measurement := ifelse(avg == "Avg(4)", "EC", measurement)]

# Replace sv with the combined table for downstream processing
sv <- sv_combined

# Continue with temp/moisture subsetting as before
temp.sv <- subset(sv, measurement == "temp")
moisture.sv <- subset(sv, measurement == "moisture")

```



# Breakup variable name so we can extract plot and sensor info
``` {r}
#Remove instrument noise
temp.sv[, num := ifelse(num<(-5), NA, num)]
temp.sv[, num := ifelse(num>40, NA, num)]

temp.sv$PP<- as.numeric(temp.sv$PP)
moisture.sv$PP <- as.numeric(moisture.sv$PP)

#setting depths
temp.sv$depth <- factor(temp.sv$depth, labels=c("5", "10", "20","30", "40","50"))
moisture.sv$depth <- factor(moisture.sv$depth, labels=c("5", "10", "20","30", "40","50"))
temp.sv$trt <- factor(temp.sv$trt,levels = c("H", "C"))
moisture.sv$trt <- factor(moisture.sv$trt, levels = c("H", "C"))
temp.sv <- temp.sv %>% filter(!is.na(num) & !is.na(depth) & !is.na(trt) & !is.na(PP))
moisture.sv <- moisture.sv %>% filter(!is.na(num) & !is.na(depth) & !is.na(trt) & !is.na(PP))


# Filter for H and C treatments separately
t_h_sv <- temp.sv %>% filter(trt == "H")
m_h_sv <- moisture.sv %>% filter(trt == "H") %>% mutate(PP = as.double(PP))
t_c_sv <- temp.sv %>% filter(trt == "C")
m_c_sv <- moisture.sv %>% filter(trt == "C") %>% mutate(PP = as.double(PP))


# Join H and C datasets based on TIMESTAMP, depth, and PP
t_diff_sv <- t_h_sv %>%
  inner_join(t_c_sv, by = c("TIMESTAMP", "depth", "PP")) %>%
  mutate(diff_H_C = num.x - num.y) %>%  # Calculate differences for each depth
  select(TIMESTAMP, depth, PP, diff_H_C)  # Select relevant column
tempdiffbymoisture <- t_diff_sv %>%
  inner_join(m_h_sv, by = c("TIMESTAMP", "depth", "PP")) %>% 
  select(TIMESTAMP, depth, PP, diff_H_C, num)
temp_filtered_Burned <- temp.sv %>% filter(PP %in% 1:4)
moisture_filtered_Burned <- moisture.sv %>% filter(PP %in% 1:4)

temp_filtered_Unburned <- temp.sv %>% filter(PP %in% 5:8)
moisture_filtered_Unburned <- moisture.sv %>% filter(PP %in% 5:8)
```

# Plots
``` {r}
#scatterplot of temp across all depths and treatments

temp.sv |>
  ggplot(aes(x = num))+
  geom_histogram()

```

``` {r} 
#HISTOGRAM FOR DISTRIBUTION OF TEMP DIFFERENCE
library(dplyr)
library(ggplot2)

tempdiffbymoisture %>%
  filter(PP == 1) %>%
  ggplot(aes(x = diff_H_C)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue", color = "black", alpha = 0.7) +
  facet_wrap(~ depth) +
  labs(
    x = "Temperature Difference (Heated - Control)",
    y = "Frequency",
    title = "Temp Diff (PP = 1)"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```


``` {r}
# KEEP
#Raw Temp data SoilVUE
ggplot(temp_filtered_Burned, aes(x = TIMESTAMP, y = num, color = trt)) +
  geom_point(shape = 16, size = 1) + 
  ylab(expression("Temperature (degree*C)")) +
  xlab("Date") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14, angle = -45, vjust = 0.5),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),  # Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic"),
    legend.title = element_blank()  # Remove legend title
  ) +
  guides(color = guide_legend(title = FALSE)) +  # Remove legend title
  facet_grid(PP ~ depth) +
  ggtitle("SoilVUE Controlled and Heated Temperature Burned Plots 1-4") +
  scale_color_manual(values = c("red","blue"))
```


# Plot 1 heated 
``` {r}
#Only for plot one because we only heated PP 1
ggplot(temp_filtered_Burned, 
       aes(x = TIMESTAMP, y = num, color = trt)) +
  geom_point(shape = 16, size = 1) + 
  ylab(expression("Temperature (degree*C)")) +
  xlab("Date") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size=14, angle = -45, vjust = 0.5),
    axis.text.y = element_text(size=14),
    axis.title.x = element_text(size=14),
    axis.title.y = element_text(size=14),
    strip.text = element_text(size=12, color="black", face="italic"),
    legend.title = element_blank()
  ) +
  guides(color = guide_legend(title = FALSE)) +
  facet_wrap(. ~ depth, ncol = 1) +  # Only facet by depth now
  ggtitle("SoilVUE Controlled and Heated Temperature Burned Plot 1") +
  scale_color_manual(values = c("red", "blue"))
```

``` {r}
#Raw Temp data SoilVUE
ggplot(temp_filtered_Unburned, aes(x = TIMESTAMP, y = num, color = trt)) +
  geom_point(shape = 16, size = 1) + 
  ylab(expression("Temperature (degree*C)")) +
  xlab("Date") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14, angle = -45, vjust = 0.5),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),  # Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic"),
    legend.title = element_blank()  # Remove legend title
  ) +
  guides(color = guide_legend(title = FALSE)) +  # Remove legend title
  facet_grid(PP ~ depth) +
  ggtitle("SoilVUE Controlled Temperature Unburned Plots 1-4") +
  scale_color_manual(values = c("red","blue"))
  #ggsave("C:/Program Files/RStudio/Tilden Park Warming/Graphs/SoilVUE Controlled Temperature Unburned Plots 1-4.jpg", width=15,height=10)
# ggsave("C:/Program Files/RStudio/Task 5 Microwarming/Plots/Research Paper/SoilVUE Controlled and Heated Temperature Plots 1-3.jpg", width=15, height=10)
```
  
  
  
```{r}
# Temperature timeseries BH1
t_h_sv |>
  filter(PP == 1) |>
  ggplot(aes(x = TIMESTAMP, y = num)) +
  geom_point(shape = 16, size = 1, color = "red") +
  ylab(expression("Temperature (degree*C)")) +
  xlab("Date") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14, angle = -45, vjust = 0.5),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),  # Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic"),
    legend.title = element_blank()  # Remove legend title
  ) +
  guides(color = guide_legend(title = FALSE)) +  # Remove legend title
  facet_grid(~ depth) +
  ggtitle("Temperature Timeseries Burned Heated Plot 1")


t_c_sv |>
  filter(PP == 1) |>
  ggplot(aes(x = TIMESTAMP, y = num)) +
  geom_point(shape = 16, size = 1, color = "blue") +
  ylab(expression("Temperature (degree*C)")) +
  xlab("Date") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14, angle = -45, vjust = 0.5),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),  # Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic"),
    legend.title = element_blank()  # Remove legend title
  ) +
  guides(color = guide_legend(title = FALSE)) +  # Remove legend title
  facet_grid(~ depth) +
  ggtitle("Temperature Timeseries Burned Control Plot 1")

```

# Summary statistics function and calculations
``` {r}
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
  return(data_sum)
}

# Convert TIMESTAMP to datetime if it's not already
temp.sv$TIMESTAMP <- as.POSIXct(temp.sv$TIMESTAMP)

# Ensure depth, trt, and PP are factors if they are categorical
temp.sv$depth <- as.factor(temp.sv$depth)
temp.sv$trt <- as.factor(temp.sv$trt)
temp.sv$PP <- as.factor(temp.sv$PP)

# Step 1: Calculate summary statistics for each group
summary_sv <- temp.sv %>%
  dplyr::group_by(TIMESTAMP, depth, trt, PP) %>%
  dplyr::summarise(
    num = mean(num, na.rm = TRUE),  # Calculate mean of 'num' within each group
  )

# Step 2: Calculate overall mean, sd, and se values per group (TIMESTAMP, depth, trt)
summary_final_sv <- summary_sv %>%
  dplyr::group_by(TIMESTAMP, depth, trt) %>%
  dplyr::summarise(
    mean = mean(num, na.rm = TRUE),  # Calculate mean of 'mean' within each group
    sd = sd(num, na.rm = TRUE),      # Calculate sd of 'num' within each group
    se = sd(num, na.rm = TRUE) / sqrt(sum(!is.na(num)))  # Calculate se of 'num' within each group
  )
```

```{r}
# Partition timeseries into daytime and nighttime windows
temp.sv$hour <- hour(temp.sv$TIMESTAMP)

# Define windows: daytime 6:00-20:00, nighttime 20:00-6:00
temp_day <- temp.sv %>%
  filter(hour >= 6 & hour < 20, !is.na(num), !is.na(depth)) %>%
  mutate(window = "Daytime (6:00-20:00)")

temp_night <- temp.sv %>%
  filter(hour >= 20 | hour < 6, !is.na(num), !is.na(depth)) %>%
  mutate(window = "Nighttime (20:00-6:00)")

# Combine
temp_windows <- rbind(temp_day, temp_night)

# Overlapping density plot by depth and treatment
ggplot(temp_windows, aes(x = num, fill = window, color = window)) +
  geom_density(alpha = 0.4, size = 1) +
  facet_grid(depth ~ trt) +
  labs(
    x = "Temperature (°C)",
    y = "Density",
    title = "Baseline Temperature Distribution: Daytime vs Nighttime",
    fill = "Time Window",
    color = "Time Window"
  ) +
  theme_bw() +
  theme(
    legend.text = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 11, face = "italic")
  )

# Optional: summary stats by window, depth, treatment
window_summary <- temp_windows %>%
  group_by(window, depth, trt) %>%
  summarise(
    mean_temp = mean(num, na.rm = TRUE),
    sd_temp = sd(num, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

window_summary
```


# Temp diff between burned and unburned 
```{r, fig.width=16, fig.height=10}

paired_diff_pp <- temp.sv %>%
  filter(PP %in% 1:8, !is.na(depth), !is.na(PP), !is.na(num)) %>%
  mutate(
    pair = if_else(as.integer(PP) <= 4L, as.integer(PP), as.integer(PP) - 4L),
    side = if_else(as.integer(PP) <= 4L, "burned", "unburned")
  ) %>%
  group_by(TIMESTAMP, depth, pair, side) %>%
  summarise(mean_num = mean(num, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = side, values_from = mean_num) %>%
  # keep only pairs where both burned and unburned exist
  filter(!is.na(burned) & !is.na(unburned)) %>%
  mutate(
    diff_burn_unburn = burned - unburned,
    pair = factor(pair),
    depth = factor(depth)
  ) %>%
  arrange(pair, depth, TIMESTAMP)

# quick paired plot (keep y-limits -10..10)
p_diff_pp <- paired_diff_pp %>%
  ggplot(aes(x = TIMESTAMP, y = diff_burn_unburn, group = pair, color = pair)) +
  geom_line(size = 0.6) +
  geom_point(size = 1) +
  facet_grid(pair ~ depth) +
  coord_cartesian(ylim = c(-10, 10)) +
  labs(x = "Date", y = "Temperature Difference (Burned - Unburned)", title = "Paired Plot Temp Diff (burned vs unburned)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

p_diff_pp

```

# histogram of temp diff between burned and unburned
```{r}
paired_diff_pp %>%
  filter(pair == 1) %>%
  ggplot(aes(x = diff_burn_unburn)) +
  geom_histogram(binwidth = 0.5, fill = "red", color = "red", alpha = 0.7) +
  facet_wrap(~ depth) +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(
    x = "Temperature Difference (Burned - Unburned)",
    y = "Frequency",
    title = "Temp Diff (PP = 1)"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

```{r}
paired_diff_pp %>%
  filter(pair == 2) %>%
  ggplot(aes(x = diff_burn_unburn)) +
  geom_histogram(binwidth = 0.5, fill = "#87AE73", color = "#87AE73", alpha = 0.7) +
  facet_wrap(~ depth) +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(
    x = "Temperature Difference (Burned - Unburned)",
    y = "Frequency",
    title = "Temp Diff (PP = 2)"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

```{r}
paired_diff_pp %>%
  filter(pair == 3) %>%
  ggplot(aes(x = diff_burn_unburn)) +
  geom_histogram(binwidth = 0.5, fill = "#54d4ebff", color = "#54d4ebff", alpha = 0.7) +
  facet_wrap(~ depth) +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(
    x = "Temperature Difference (Burned - Unburned)",
    y = "Frequency",
    title = "Temp Diff (PP = 3)"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

```{r}
paired_diff_pp %>%
  filter(pair == 4) %>%
  ggplot(aes(x = diff_burn_unburn)) +
  geom_histogram(binwidth = 0.5, fill = "#ab71ecff", color = "#ab71ecff", alpha = 0.7) +
  facet_wrap(~ depth) +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(
    x = "Temperature Difference (Burned - Unburned)",
    y = "Frequency",
    title = "Temp Diff (PP = 4)"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

# Temp Difference between Heated and Unheated
```{r, fig.width=16, fig.height=10}
# Compute paired differences between H and C treatments within the same PP
# pivot so H and C are side-by-side and compute difference
paired_diff_df <- temp.sv %>%
  filter(!is.na(PP) & !is.na(depth) & !is.na(trt)) %>%
  group_by(TIMESTAMP, depth, PP, trt) %>%
  summarise(mean_num = mean(num, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = trt, values_from = mean_num, names_prefix = "mean_") %>%
  # require both mean_H and mean_C to be present for a paired diff
  filter(!is.na(mean_H) & !is.na(mean_C)) %>%
  mutate(
    diff_num = mean_H - mean_C,
    PP = factor(PP),
    depth = factor(depth)
  )

# quick plot (paired diffs)
p_diff <- paired_diff_df %>%
  arrange(PP, depth, TIMESTAMP) %>%
  ggplot(aes(x = TIMESTAMP, y = diff_num, group = PP, color = PP)) +
  geom_line(size = 0.6) +
  geom_point(size = 1) +
  facet_grid(PP ~ depth) +
  coord_cartesian(ylim = c(-10, 10)) +
  labs(x = "Date", y = "Temperature Difference (H - C)", title = "Paired Plot Temperature Difference") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

p_diff
```


```{r}
#MOISTURE DIFF BETWEEN BURNED AND UNBURNED
# Step 1: Summarize Burned data
newburned <- moisture_filtered_Burned %>% filter(trt != "H")
burned_summary <- newburned %>%
  group_by(TIMESTAMP, depth) %>%
  summarize(mean_num_burned = mean(num, na.rm = TRUE), .groups = "drop")

# Step 2: Summarize Unburned data
unburned_summary <- moisture_filtered_Unburned %>%
  group_by(TIMESTAMP, depth) %>%
  summarize(mean_num_unburned = mean(num, na.rm = TRUE), .groups = "drop")

# Step 3: Merge the two summaries by TIMESTAMP and Depth (inner join to get shared ones)
moist_compare <- inner_join(burned_summary, unburned_summary, by = c("TIMESTAMP", "depth"))

# Step 2: Reshape from wide to long format
long_df <- moist_compare %>%
  pivot_longer(cols = c(mean_num_burned, mean_num_unburned),
               names_to = "condition",
               values_to = "mean_moisture") %>%
  mutate(condition = recode(condition,
                            mean_num_burned = "Burned",
                            mean_num_unburned = "Unburned"))

# Step 3: Plot burned and unburned on same plot
ggplot(long_df, aes(x = TIMESTAMP, y = mean_moisture, color = condition)) +
  geom_line(size = 1.5) +
  facet_wrap(~ depth, scales = "free_y") +
  labs(
    x = "Timestamp",
    y = "Mean Moisture",
    color = "Condition"
  ) +
  theme_minimal() +
  theme(
    legend.text = element_text(size=20),
    legend.title = element_text(size=20),
    axis.text.x = element_text(size=20, angle = -45, vjust = 0.5),
    axis.text.y = element_text(size=20),
    axis.title.x = element_text(size=20),
    axis.title.y = element_text(size=20),
    strip.text = element_text(size=18, color="black", face="italic")
  )

```

``` {r}
#standard deviation all depths
ggplot(summary_final_sv, aes(x = TIMESTAMP, y = mean, colour = trt)) +
  geom_point(size = 0.5) +
  geom_ribbon(aes(ymin = mean - sd, ymax = mean + sd, fill = trt), alpha = 0.5, colour = NA) +
  scale_color_manual(values = c("C" = "blue", "H" = "red")) +
  scale_fill_manual(values = c("C" = "blue", "H" = "red")) +
  ylab("Mean Temp (degree*C)") +
  xlab("Date") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),# Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic")
  ) +
  facet_grid(~depth) +
  ggtitle("SoilVUE Mean Temperature & Standard Deviation Plots 1-3") +
  scale_x_datetime(breaks = "8 days", date_labels = "%b-%d") +
  theme(axis.text.x = element_text(angle = -45)) +
  labs(color = NULL, fill = NULL)
```

``` {r}
#KEEP 
#standard deviation plots 30 cm depth
summary_final_sv_30cm <- subset(summary_final_sv, depth == '30')

ggplot(summary_final_sv_30cm, aes(x = TIMESTAMP, y = mean, colour = trt)) +
  geom_point(size = 0.5) +
  geom_ribbon(aes(ymin = mean - sd, ymax = mean + sd, fill = trt), alpha = 0.5, colour = NA) +
  scale_color_manual(values = c("C" = "blue", "H" = "red")) +
  scale_fill_manual(values = c("C" = "blue", "H" = "red")) +
  ylab("Temperature (degree*C)") +
  xlab("Date") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),# Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic")
  ) +
  ggtitle("SoilVUE Temperature Standard Deviation Plot for 30 cm Depth") +
  scale_x_datetime(breaks = "8 days", date_labels = "%b-%d") +
  theme(axis.text.x = element_text(angle = -45)) +
  labs(color = NULL, fill = NULL)
```



``` {r}
#Keep
# moisture H and C (raw)
ggplot(moisture.sv, aes(x=TIMESTAMP, y=num, colour=trt)) +
  geom_point(shape=16, size=1) + 
  ylab(expression("Moisture Content")) +
  xlab("Date") +
  scale_color_manual(values = c("C" = "blue", "H" = "red")) +
  ggtitle("SoilVUE Moisture Plots 1-4") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14, angle = -45, vjust = 0.5),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),# Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic")
  ) +
  guides(colour = guide_legend(title = NULL)) +
  facet_grid(PP~depth)
  
  #ggsave("C:/Program Files/RStudio/Tilden Park Warming/Graphs/SoilVUE Moisture Plots 1-4.jpg", width=15,height=10)
# ggsave("C:/Program Files/RStudio/Task 5 Microwarming/Plots/Research Paper/SoilVUE Moisture Plots 1-3.pdf.jpg", width=15, height=10)

library(ggplot2)

# Filter the data for PP 3
moisture.sv_PP3 <- subset(moisture.sv, PP == 3)
```

``` {r}
#Keep 
# Create raw data plot 3
ggplot(moisture.sv_PP3, aes(x=TIMESTAMP, y=num, colour=trt)) +
  geom_point(shape=5, size=0.5) + 
  ylab(expression("Moisture Content")) +
  xlab("Date") +
  scale_color_manual(values = c("C" = "blue", "H" = "red")) +
  #  scale_x_datetime(
  #   date_breaks = "10 days",       # Set break intervals to 20 days
  #    date_labels = "%d %b"       # Format the date labels
  # ) +
  ggtitle("SoilVUE Moisture Plot 3") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 12, angle = -45, vjust = 0.5),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),# Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic")
  ) +
  guides(colour = guide_legend(title = NULL)) +
  facet_grid(. ~ depth)
# ggsave("C:/Program Files/RStudio/Task 5 Microwarming/Plots/SoilVUE Moisture Plot 3.pdf", width=10, height=5)
# ggsave("C:/Program Files/RStudio/Task 5 Microwarming/Plots/SoilVUE Moisture Plot 3.jpg", width=10, height=5)


moisture_num <- moisture.sv$num

names(temp.sv) <- c("TIMESTAMP", "variable", "temp_num", "plot", "depth", "avg", "PP", "trt", "measurement")

temp_moisture_sv <- cbind(temp.sv, moisture_num = moisture.sv$num)


#Keep 
# Temp vs moisture 
ggplot(temp_moisture_sv, aes(x = temp_num, y = moisture_num, colour = trt)) +
  geom_point(shape = 1, size = 0.75) +
  ylab(expression("Soil Moisture")) +
  xlab("Temperature (degree*C)") +
  labs(title = "SoilVUE Temperature vs Moisture Plots 1-3") +
  theme_bw() +
  facet_grid(PP ~ depth) +
  scale_color_manual(values = c("C" = "blue", "H" = "red"))
# ggsave("C:/Program Files/RStudio/Task 5 Microwarming/Plots/SoilVUE Temperature vs Moisture Plots 1-3.pdf", width=15,height=10,dpi=300)


summary.moisture2 <- moisture.sv %>%
  dplyr::group_by(TIMESTAMP,depth, trt) %>%
  dplyr::summarise(
    N=length(num),
    mean= mean(num, na.rm=T),
    sd=sd(num, na.rm=T),
    se=sd(num, na.rm=T)/sqrt(sum(!is.na(num))))


#KEEP
#standard deviation moisture 
ggplot(summary.moisture2, aes(x = TIMESTAMP, y = mean, colour = trt)) +
  geom_point(size = 1) +
  geom_ribbon(aes(ymin = mean - sd, ymax = mean + sd, fill = trt), alpha = 0.5, colour = NA) +
  scale_color_manual(values = c("C" = "blue", "H" = "red")) +
  scale_fill_manual(values = c("C" = "blue", "H" = "red")) +
  ylab("Moisture Content") +
  xlab("Date") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),# Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic")
  ) +
  facet_grid(~depth) +
  ggtitle("SoilVUE Moisture Mean & Standard Deviation Plots 1-3") +
  scale_x_datetime(breaks = "8 days", date_labels = "%b-%d") +
  theme(axis.text.x = element_text(angle = -45)) +
  labs(color = NULL, fill = NULL)



#SoilVUE Electrical Conductivity
EC.sv<- subset(sv, measurement=="EC")


EC.sv$PP<- as.numeric(EC.sv$PP)

unique(EC.sv$depth)
EC.sv$depth <- factor(EC.sv$depth, labels=c("5 cm", "10 cm", "20 cm","30 cm", "40 cm","50 cm"))


#KEEP
# EC H and C (raw)
ggplot(EC.sv, aes(x=TIMESTAMP, y=num, colour=trt)) +
  geom_point(size=0.5) + 
  ylab(expression("Electrical Conductivity (dS/m)")) +
  xlab("Date") +
  scale_color_manual(values = c("C" = "blue", "H" = "red")) +
  scale_x_datetime(
    date_breaks = "20 days",       # Set break intervals to 20 days
    date_labels = "%d %b"       # Format the date labels
  ) +
  ggtitle("SoilVUE Electrical Conductivity Plots 1-3") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),# Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic")
  ) +
  guides(colour = guide_legend(title = NULL)) +
  facet_grid(PP~depth)


summary.EC2 <- EC.sv %>%
  dplyr::group_by(TIMESTAMP,depth, trt) %>%
  dplyr::summarise(
    N=length(num),
    mean= mean(num, na.rm=T),
    sd=sd(num, na.rm=T),
    se=sd(num, na.rm=T)/sqrt(sum(!is.na(num))))


#KEEP
#Standard deviation all depths
ggplot(summary.EC2, aes(x = TIMESTAMP, y = mean, colour = trt)) +
  geom_point(size = 1) +
  geom_ribbon(aes(ymin = mean - sd, ymax = mean + sd, fill = trt), alpha = 0.5, colour = NA) +
  scale_color_manual(values = c("C" = "blue", "H" = "red")) +
  scale_fill_manual(values = c("C" = "blue", "H" = "red")) +
  ylab("Electrical Conductivity (dS/m)") +
  xlab("Date") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),# Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic")
  ) +
  facet_grid(~depth) +
  ggtitle("SoilVUE Electrical Conductivity Mean & Standard Deviation Plots 1-3") +
  scale_x_datetime(breaks = "8 days", date_labels = "%b-%d") +
  theme(axis.text.x = element_text(angle = -45)) +
  labs(color = NULL, fill = NULL)


#SoilVUE Permitivity
PER.sv<- subset(sv, measurement=="permitivity")


PER.sv$PP<- as.numeric(PER.sv$PP)

unique(PER.sv$depth)
PER.sv$depth <- factor(PER.sv$depth, labels=c("5 cm", "10 cm", "20 cm","30 cm", "40 cm","50 cm"))


#KEEP
# EC H and C (raw)
ggplot(PER.sv, aes(x=TIMESTAMP, y=num, colour=trt)) +
  geom_point(size=0.5) + 
  ylab(expression("Permittivity (ɛ)")) +
  xlab("Date") +
  scale_color_manual(values = c("C" = "blue", "H" = "red")) +
  scale_x_datetime(
    date_breaks = "20 days",       # Set break intervals to 20 days
    date_labels = "%d %b"       # Format the date labels
  ) +
  ggtitle("SoilVUE Permittivity Plots 1-3") +
  theme_bw() +
  theme(
    legend.text = element_text(size=14),
    axis.text.x = element_text(size= 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size=14),  # Increase x-axis label font size
    axis.title.y = element_text(size=14),# Increase y-axis label font size
    strip.text = element_text(size = 12, color = "black", face = "italic")
  ) +
  guides(colour = guide_legend(title = NULL)) +
  facet_grid(PP~depth)
#  ggsave("C:/Program Files/RStudio/Task 5 Microwarming/Plots/SoilVUE Moisture Plots 1-3.pdf", width=15, height=10)
 # ggsave("C:/Program Files/RStudio/Task 5 Microwarming/Plots/Research Paper/SoilVUE Permittivity Plots 1-3.jpg", width=15, height=10)
```


