---
title: "SVtilden_Code_emmaedit"
format: pdf
---

# Required packages to run script. If not on computer download and install.
``` {r}
library(ggplot2)
library(lubridate)
library(reshape2)
library(data.table)
library(dplyr)
library(gridExtra)
library(zoo)
library(plotly)
library(gridExtra)
library(data.table)
library(slider)
library(tidyverse)
```



# Reading in pre-cleaned CSV and converting to data table
``` {r}

# Reading in pre-cleaned CSV and converting to data table
setwd("C:/Users/emmam/Downloads/senior thesis/DataForWorking")

# Read the full CSV
data <- read.csv("20260406_Tilden_Data_foremma.csv", header = TRUE)
data <- data.table(data)

# --- FIX: Parse TIMESTAMP immediately after reading ---
data$TIMESTAMP <- as.POSIXct(
  strptime(data$TIMESTAMP, format = "%m/%d/%Y %H:%M", tz = "US/Pacific")
)

# Remove duplicates from data
data <- data[!duplicated(data), ]

# Order file based on timestamp
data <- data[order(TIMESTAMP), ]

# Select SoilVUE columns
SVburnt <- cbind(data[, c(1, 69:356)])

# Melt to long format
sv <- data.table::melt(SVburnt, 
                       id.vars = "TIMESTAMP", 
                       value.name = "num")

# Parse the variable column into plot, depth, and avg
sv[, c("rest", "plot", "depth", "avg") := tstrsplit(variable, "_", fixed = TRUE)]
sv[, variable := NULL]


```


# assigning treatments etc

``` {r}
# Now assign PP, trt, measurement using the parsed columns
sv[, PP := as.character(NA)]
sv[, PP := fifelse(plot == "BH1" | plot == "BC1", "1", PP)]
sv[, PP := fifelse(plot == "BH2" | plot == "BC2", "2", PP)]
sv[, PP := fifelse(plot == "BH3" | plot == "BC3", "3", PP)]
sv[, PP := fifelse(plot == "BH4" | plot == "BC4", "4", PP)]
sv[, PP := fifelse(plot == "NB1", "5", PP)]
sv[, PP := fifelse(plot == "NB2", "6", PP)]
sv[, PP := fifelse(plot == "NB3", "7", PP)]
sv[, PP := fifelse(plot == "NB4", "8", PP)]

sv[, trt := as.character(NA)]
sv[, trt := fifelse(plot %in% c("BH1","BH2","BH3","BH4"), "H", trt)]
sv[, trt := fifelse(plot %in% c("BC1","BC2","BC3","BC4"), "C", trt)]
sv[, trt := fifelse(plot %in% c("NB1","NB2","NB3","NB4"), "H", trt)]


sv[, measurement := as.character(NA)]
sv[, measurement := fifelse(avg == "Avg.1.", "moisture", measurement)]
sv[, measurement := fifelse(avg == "Avg.2.", "permitivity", measurement)]
sv[, measurement := fifelse(avg == "Avg.3.", "temp", measurement)]
sv[, measurement := fifelse(avg == "Avg.4.", "EC", measurement)]

# Subset by measurement type
temp.sv <- sv[measurement == "temp"]
moisture.sv <- sv[measurement == "moisture"]
ec.sv <- sv[measurement == "EC"]

# Inspect to confirm parsing worked
head(sv[, .(TIMESTAMP, plot, depth, avg, measurement, num)])
```


# Breakup variable name so we can extract plot and sensor info
``` {r}
#Remove instrument noise
temp.sv[, num := ifelse(num<(-5), NA, num)]
temp.sv[, num := ifelse(num>40, NA, num)]
ec.sv[, num := ifelse(num==(-9999), NA, num)] #may 2025 powershutoff all have this value as placeholder, remove this data

temp.sv$PP<- as.numeric(temp.sv$PP)
moisture.sv$PP <- as.numeric(moisture.sv$PP)
ec.sv$PP <- as.numeric(ec.sv$PP)

#setting depths
temp.sv$depth <- factor(temp.sv$depth, labels=c("5", "10", "20","30", "40","50"))
moisture.sv$depth <- factor(moisture.sv$depth, labels=c("5", "10", "20","30", "40","50"))
ec.sv$depth <- factor(ec.sv$depth, labels=c("5", "10", "20","30", "40","50"))

#setting treatments
temp.sv$trt <- factor(temp.sv$trt,levels = c("H", "C"))
moisture.sv$trt <- factor(moisture.sv$trt, levels = c("H", "C"))
ec.sv$trt <- factor(ec.sv$trt, levels = c("H", "C"))

temp.sv <- temp.sv |>
  mutate(burnstatus = case_when(
    plot %in% c("BH1", "BH2", "BH3", "BH4") ~ "Heated",
    plot %in% c("BC1", "BC2", "BC3", "BC4") ~ "Burned",
    plot %in% c("NB1", "NB2", "NB3", "NB4") ~ "Unburned",
    TRUE ~ NA_character_
  ))

temp.sv$burnstatus <- factor(temp.sv$burnstatus, levels = c("Heated", "Burned", "Unburned"))

#remove missing data
temp.sv <- temp.sv %>% filter(!is.na(num) & !is.na(depth) & !is.na(trt) & !is.na(PP))
moisture.sv <- moisture.sv %>% filter(!is.na(num) & !is.na(depth) & !is.na(trt) & !is.na(PP))
ec.sv <- ec.sv %>% filter(!is.na(num) & !is.na(depth) & !is.na(trt) & !is.na(PP))

```

# plot-level differences
```{r}
temp.sv |>
  filter(plot == "NB1") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Nonburned Plot 1"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

temp.sv |>
  filter(plot == "NB2") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Nonburned Plot 2")+
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

temp.sv |>
  filter(plot == "NB3") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Nonburned Plot 3")+
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

temp.sv |>
  filter(plot == "NB4") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Nonburned Plot 4")+
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

# plot-level differences with all nonburned plots together
```{r}
# Plot all nonburned plots with depths in rows, plots in columns
temp.sv %>%
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4")) %>%
  ggplot(aes(x = num)) +
  geom_histogram(binwidth = 0.5, fill = "darkgreen", color = "darkgreen")+
  facet_grid(depth ~ plot) +
  labs(
    x = "Temperature (°C)",
    y = "Frequency",
    title = "Nonburned Plots Temperature Distribution by Depth"
  ) +
  theme_bw(base_size = 30)
```

# Figure 3: Unburned by Depth Temp Dists
```{r}
# Overlapping density plot by depth and treatment

temp.sv |>
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4")) |>
  ggplot(aes(x = num, fill = plot, color = plot)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ depth) +
  labs(
    x = "Temperature (°C)",
    y = "Density",
    title = "Nonburned Plots Temperature Distribution",
    fill = "Plot",
    color = "Plot"
  ) +
  theme_bw(base_size = 30)
```


# Figure 1: Temperature distribution in unburned plots all depths
```{r}
temp.sv |> 
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4")) |>
  ggplot(aes(x = num, fill = plot, color = plot)) +
  geom_density(alpha = 0.4) +
  labs(
    x = "Temperature (°C)",
    y = "Density",
    title = "Nonburned Plots Temperature Distribution",
    fill = "Plot",
    color = "Plot"
  ) +
  theme_bw(base_size = 50)

```

# Average Nonburnt 
```{r}
nonburnt_avg <- temp.sv %>%
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4")) %>%
  group_by(TIMESTAMP, depth) %>%
  summarise(mean_num = mean(num, na.rm = TRUE), .groups = "drop") 

nonburnt_avg |>
  ggplot(aes(x = mean_num)) +
  geom_histogram(binwidth = 0.5, fill = "darkgreen", color = "darkgreen") +
  facet_wrap(~ depth) +
  labs(
    x = "Temperature (°C)",
    y = "Frequency",
    title = "Nonburned Plots Average Temperature Distribution by Depth"
  ) +
  theme_bw() +
  theme(
    legend.text = element_text(size = 12),
    axis.text.x = element_text(size = 10, angle = -45, hjust = 0),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 11, face = "italic")
  )

```

# burned plot-level differences

```{r}
temp.sv |>
  filter(plot == "BC1") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Burned Control Plot 1"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

temp.sv |>
  filter(plot == "BC2") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Burned Control Plot 2")+
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

temp.sv |>
  filter(plot == "BC3") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Burned Control Plot 3")+
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

temp.sv |>
  filter(plot == "BC4") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Burned Control Plot 4")+
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

temp.sv |>
  filter(plot == "BH2") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Burned Heat Plot 2"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

temp.sv |>
  filter(plot == "BH3") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Burned Heat Plot 3"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

temp.sv |>
  filter(plot == "BH4") |>
  ggplot(aes(x = num))+
  geom_histogram(binwidth = 0.5)+
  facet_wrap(~ depth) +
  labs(
    x = "Temperature",
    y = "Frequency",
    title = "Temp of Burned Heat Plot 4"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "italic"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )  
```

# burned plots together by depth and plot in columns
```{r}
temp.sv %>%
  filter(plot %in% c("BC1", "BC2", "BC3", "BC4")) %>%
  ggplot(aes(x = num)) +
  geom_histogram(binwidth = 0.5, fill = "darkblue", color = "darkblue")+
  facet_grid(depth ~ plot) +
  labs(
    x = "Temperature (°C)",
    y = "Frequency",
    title = "Burned Plots Temperature Distribution by Depth"
  ) +
  theme_bw(base_size = 30)
```

# Figure 2: All temperature data in burned plots at all depths
```{r}
temp.sv |>
  filter(plot %in% c("BC1", "BC2", "BC3", "BC4")) |>
  ggplot(aes(x = num, fill = plot, color = plot)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ depth) +
  labs(
    x = "Temperature (°C)",
    y = "Density",
    title = "Burned Plots Temperature Distribution",
    fill = "Plot",
    color = "Plot"
  ) +
  theme_bw(base_size = 30)
```

# Figure 4: Temperature distribution in burned plots by depth
```{r}
temp.sv |>
  filter(plot %in% c("BC1", "BC2", "BC3", "BC4")) |>
  ggplot(aes(x = num, fill = plot, color = plot)) +
  geom_density(alpha = 0.4) +
  labs(
    x = "Temperature (°C)",
    y = "Density",
    title = "Burned Plots Temperature Distribution",
    fill = "Plot",
    color = "Plot"
  ) +
  theme_bw(base_size = 50)


```

# Average burned plots by depth
```{r}
burnt_avg <- temp.sv %>%
  filter(plot %in% c("BC1", "BC2", "BC3", "BC4")) %>%
  group_by(TIMESTAMP, depth) %>%
  summarise(mean_num = mean(num, na.rm = TRUE), .groups = "drop")

burnt_avg |>
  ggplot(aes(x = mean_num)) +
  geom_histogram(binwidth = 0.5, fill = "darkblue", color = "darkblue") +
  facet_wrap(~ depth) +
  labs(
    x = "Temperature (°C)",
    y = "Frequency",
    title = "Burned Plots Average Temperature Distribution by Depth"
  ) +
  theme_bw()

```


# Figure 5: Average temperature distribution in burned and unburned plots by depth
```{r}

temp.sv |>
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4", "BC1", "BC2", "BC3", "BC4")) |>
  ggplot(aes(x = num, fill = burnstatus, color = burnstatus)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ depth) +
  labs(
    x = "Temperature (°C)",
    y = "Density",
    title = "Temperature Distribution: Burned vs Nonburned",
    fill = "Burn Status",
    color = "Burn Status"
  ) +
  theme_bw(base_size = 20)
```

# Figure 6: Average temperature in burned and unburned plots by depth through time
```{r}
temp.sv |>
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4", "BC1", "BC2", "BC3", "BC4")) |>
  ggplot(aes(x = TIMESTAMP, y = num, color = burnstatus)) +
  geom_point(size = 0.5, alpha = 0.5) +
  facet_grid(. ~ depth) +
  scale_x_datetime(
    date_breaks = "3 months",
    date_labels = "%b %Y"
  ) +
  labs(
    x = "Date",
    y = "Temperature (°C)",
    title = "Temperature Over Time by Burn Status",
    color = "Treatment"
  ) +
  theme_bw(base_size = 20)
```

# Figure 7: Difference in average temperature for burned and unburned plots by depth through time
```{r}
# Figure 7: Average Burned - Average Nonburned
temp_diff <- temp.sv %>%
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4", "BC1", "BC2", "BC3", "BC4")) %>%
  group_by(TIMESTAMP, depth, burnstatus) %>%
  summarise(mean_num = mean(num, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = burnstatus, values_from = mean_num) %>%
  filter(!is.na(Burned) & !is.na(Unburned)) %>%
  mutate(diff_burn_nonburn = Burned - Unburned)

# quick paired plot (keep y-limits -10..10)
p_diff <- temp_diff %>%
  ggplot(aes(x = TIMESTAMP, y = diff_burn_nonburn, group = depth, color = depth)) +
  geom_line(size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_point(size = 1) +
  facet_grid(depth ~ .) +
  coord_cartesian(ylim = c(-10, 10)) +
  labs(x = "Date", y = "Temperature Difference (Burned - Nonburned)", title = "Temp Diff (Burned vs Nonburned)") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = -45, hjust = 0))

p_diff
```

# Figure 8: Average temp difference distributions by depth
```{r}
temp_diff |>
  ggplot(aes(x = diff_burn_nonburn)) +
  geom_density(fill = "purple", color = "purple") +
  facet_wrap(~ depth) +
  labs(
    x = "Temperature Difference (Burned - Nonburned)",
    y = "Frequency",
    title = "Temperature Difference Distribution (Burned vs Nonburned)"
  ) +
  theme_bw(base_size = 50)
```

# Water Year Analysis
```{r}

temp.sv <- temp.sv |>
  mutate(water_year = ifelse(TIMESTAMP >= as.POSIXct("2024-11-12") & TIMESTAMP < as.POSIXct("2025-11-12"), "Water Year 1", 
                           ifelse(TIMESTAMP >= as.POSIXct("2025-11-12") & TIMESTAMP < as.POSIXct("2026-11-12"), "Water Year 2", NA))) |>
  filter(!is.na(water_year))

temp.sv$water_year <- factor(temp.sv$water_year, levels = c("Water Year 1", "Water Year 2"))

# Create aligned water year data for overlaying
temp.sv_wy <- temp.sv %>%
  filter(!is.na(water_year)) %>%
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4", "BC1", "BC2", "BC3", "BC4")) %>%
  mutate(
    # Extract month and day-of-year for alignment
    month_num = month(TIMESTAMP),
    # Convert to "water year month" (Nov=11, Dec=12, Jan=1, ..., Apr=4 becomes 1-5)
    wy_month = ifelse(month_num >= 11, month_num - 10, month_num + 2)
  ) %>%
  filter(wy_month >= 1 & wy_month <= 6) %>%  # Keep only Nov-Apr
  group_by(water_year, depth, burnstatus, wy_month) %>%
  summarise(mean_temp = mean(num, na.rm = TRUE), .groups = "drop")

# Plot both water years overlaid
ggplot(temp.sv_wy, aes(x = wy_month, y = mean_temp, color = water_year, linetype = water_year)) +
  geom_line(linewidth = 3) +
  geom_point(size = 3) +
  facet_grid(depth ~ burnstatus) +
  scale_x_continuous(
    breaks = 1:6,
    labels = c("Nov", "Dec", "Jan", "Feb", "Mar", "Apr")
  ) +
  labs(
    x = "Month",
    y = "Temperature (°C)",
    title = "Temperature Comparison: Water Year 1 vs Water Year 2",
    color = "Water Year",
    linetype = "Water Year"
  ) +
  theme_bw(base_size = 30) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```

```{r}
# Water Year Analysis - Temperature Difference (Burned - Nonburned)
temp_wy <- temp.sv |>
  mutate(water_year = ifelse(TIMESTAMP >= as.POSIXct("2024-11-12") & TIMESTAMP < as.POSIXct("2025-04-06"), "Water Year 1", 
                           ifelse(TIMESTAMP >= as.POSIXct("2025-11-12") & TIMESTAMP < as.POSIXct("2026-04-06"), "Water Year 2", NA))) |>
  filter(!is.na(water_year))

temp_wy$water_year <- factor(temp_wy$water_year, levels = c("Water Year 1", "Water Year 2"))

# Create aligned water year data with burned-nonburned differences
temp_wy_diff <- temp_wy %>%
  filter(!is.na(water_year)) %>%
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4", "BC1", "BC2", "BC3", "BC4")) %>%
  group_by(TIMESTAMP, depth, water_year, burnstatus) %>%
  summarise(mean_temp = mean(num, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = burnstatus, values_from = mean_temp) %>%
  filter(!is.na(Burned) & !is.na(Unburned)) %>%
  mutate(
    temp_diff = Burned - Unburned,
    month = month(TIMESTAMP)) %>%
  filter(month >= 11 | month <= 4) # Keep only Nov-Apr

# Plot both water years overlaid with 0 baseline
temp_wy_diff |>
  ggplot(aes(x = month, y = temp_diff, color = water_year, linetype = water_year)) +
  geom_line(linewidth = 3) +
  geom_point(size = 3) +
  facet_grid(depth ~ .) +
  scale_x_continuous(
    breaks = c(11, 12, 1, 2, 3, 4),
    labels = c("Nov", "Dec", "Jan", "Feb", "Mar", "Apr")
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(
    x = "Month",
    y = "Temperature Difference (Burned - Nonburned)",
    title = "Temperature Difference Comparison: Water Year 1 vs Water Year 2",
    color = "Water Year",
    linetype = "Water Year"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")

```

# Summary statistics function and calculations
``` {r}
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
  return(data_sum)
}

# Convert TIMESTAMP to datetime if it's not already
temp.sv$TIMESTAMP <- as.POSIXct(temp.sv$TIMESTAMP)

# Ensure depth, trt, and PP are factors if they are categorical
temp.sv$depth <- as.factor(temp.sv$depth)
temp.sv$trt <- as.factor(temp.sv$trt)
temp.sv$PP <- as.factor(temp.sv$PP)


# Step 1: Calculate summary statistics for each group
summary_sv <- temp.sv %>%
  dplyr::group_by(TIMESTAMP, depth, trt, PP) %>%
  dplyr::summarise(
    num = mean(num, na.rm = TRUE),  # Calculate mean of 'num' within each group
  )

# Step 2: Calculate overall mean, sd, and se values per group (TIMESTAMP, depth, trt)
summary_final_sv <- summary_sv %>%
  dplyr::group_by(TIMESTAMP, depth, trt) %>%
  dplyr::summarise(
    mean = mean(num, na.rm = TRUE),  # Calculate mean of 'mean' within each group
    sd = sd(num, na.rm = TRUE),      # Calculate sd of 'num' within each group
    se = sd(num, na.rm = TRUE) / sqrt(sum(!is.na(num)))  # Calculate se of 'num' within each group
  )
```

# seperating into day and night windows
```{r}
# Partition timeseries into daytime and nighttime windows
temp.sv$hour <- hour(temp.sv$TIMESTAMP)

# Define windows: daytime 6:00-20:00, nighttime 20:00-6:00
temp_day <- temp.sv %>%
  filter(hour >= 6 & hour < 20, !is.na(num), !is.na(depth)) %>%
  mutate(window = "Daytime (6:00-20:00)")

temp_night <- temp.sv %>%
  filter(hour >= 20 | hour < 6, !is.na(num), !is.na(depth)) %>%
  mutate(window = "Nighttime (20:00-6:00)")

# Combine
temp_windows <- rbind(temp_day, temp_night)
temp_windows <- temp_windows |>
  mutate(burnstatus = ifelse(plot %in% c("NB1", "NB2", "NB3", "NB4"), "Unburned", "Burned"))

# Overlapping density plot by depth and treatment
ggplot(temp_windows, aes(x = num, fill = window, color = window)) +
  geom_density(alpha = 0.4, color= NA) +
  facet_grid(depth ~ burnstatus) +
  labs(
    x = "Temperature (°C)",
    y = "Density",
    title = "Baseline Temperature Distribution: Daytime vs Nighttime",
    fill = "Time Window",
    color = "Time Window"
  ) +
  theme_bw() +
  theme(
    legend.text = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 11, face = "italic")
  )

# Optional: summary stats by window, depth, treatment
window_summary <- temp_windows %>%
  group_by(window, depth, burnstatus) %>%
  summarise(
    mean_temp = mean(num, na.rm = TRUE),
    sd_temp = sd(num, na.rm = TRUE),
    max_temp = max(num, na.rm =TRUE),
    n = n(),
    .groups = "drop"
  )

window_summary |>
  print(n=24)
```


# Moisture Analysis 

```{r}
moisture.sv <- moisture.sv |>
  mutate(burnstatus = case_when(
    plot %in% c("BH1", "BH2", "BH3", "BH4") ~ "Heated",
    plot %in% c("BC1", "BC2", "BC3", "BC4") ~ "Burned",
    plot %in% c("NB1", "NB2", "NB3", "NB4") ~ "Unburned",
    TRUE ~ NA_character_
  ))

moisture.sv$burnstatus <- factor(moisture.sv$burnstatus, levels = c("Heated", "Burned", "Unburned"))

moisture.sv |>
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4", "BC1", "BC2", "BC3", "BC4")) |>
  ggplot(aes(x = num, fill = burnstatus, color = burnstatus)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ depth) +
  labs(
    x = "Moisture Content",
    y = "Density",
    title = "SoilVUE Moisture Distribution by Depth and Treatment",
    fill = "Treatment",
    color = "Treatment"
  ) +
  theme_bw() +
  theme(
    legend.text = element_text(size = 12),
    axis.text.x = element_text(size = 10, angle = -45, hjust = 0),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 11, face = "italic")
  )
```


```{r}
#figure 1: nonburned moisture distribution all depths
moisture.sv |> 
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4")) |>
  ggplot(aes(x = num, fill = plot, color = plot)) +
  geom_density(alpha = 0.4) +
  labs(
    x = "Moisture Content",
    y = "Density",
    title = "Nonburned Plots Moisture Distribution",
    fill = "Plot",
    color = "Plot"
  ) +
  theme_bw(base_size = 50)
```


```{r}
# figure 2: burned moisture distribution all depths
moisture.sv |>
  filter(plot %in% c("BC1", "BC2", "BC3", "BC4")) |>
  ggplot(aes(x = num, fill = plot, color = plot)) +
  geom_density(alpha = 0.4) +
  labs(
    x = "Moisture Content",
    y = "Density",
    title = "Burned Plots Moisture Distribution",
    fill = "Plot",
    color = "Plot"
  ) +
  theme_bw(base_size = 50)
```

```{r}
#figure 3: nonburned moisture distribution by depth
moisture.sv |>
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4")) |>
  ggplot(aes(x = num, fill = plot, color = plot)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ depth) +
  labs(
    x = "Moisture Content",
    y = "Density",
    title = "Nonburned Moisture Distribution",
    fill = "Plot",
    color = "Plot"
  ) +
  theme_bw(base_size = 30)
```


```{r}
#figure 4: burned moisture distribution by depth

moisture.sv |>
  filter(plot %in% c("BC1", "BC2", "BC3", "BC4")) |>
  ggplot(aes(x = num, fill = plot, color = plot)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ depth) +
  labs(
    x = "Moisture Content",
    y = "Density",
    title = "Burned Moisture Distribution",
    fill = "Plot",
    color = "Plot"
  ) +
  theme_bw(base_size = 30)
```


```{r}
# Figure 5. Average moisture distribution in burned and unburned plots by depth
moisture.sv |>
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4", "BC1", "BC2", "BC3", "BC4")) |>
  ggplot(aes(x = num, fill = burnstatus, color = burnstatus)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ depth) +
  labs(
    x = "Moisture Content",
    y = "Density",
    title = "Moisture Distribution: Burned vs Nonburned",
    fill = "Burn Status",
    color = "Burn Status"
  ) +
  theme_bw(base_size = 20)

```


```{r}
# Figure 6. Average moisture in burned and unburned plots by depth through time.  

moisture.sv |>
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4", "BC1", "BC2", "BC3", "BC4")) |>
  ggplot(aes(x = TIMESTAMP, y = num, color = burnstatus)) +
  geom_point(size = 0.5, alpha = 0.5) +
  facet_grid(. ~ depth) +
  scale_x_datetime(
    date_breaks = "3 months",
    date_labels = "%b %Y"
  ) +
  labs(
    x = "Date",
    y = "Moisture Content",
    title = "Moisture Over Time by Burn Status",
    color = "Treatment"
  ) +
  theme_bw(base_size = 20)
```


```{r}
#Figure 7. Difference in average temperature for burned and unburned plots by depth through time

moisture_diff <- moisture.sv %>%
  filter(plot %in% c("NB1", "NB2", "NB3", "NB4", "BC1", "BC2", "BC3", "BC4")) %>%
  group_by(TIMESTAMP, depth, burnstatus) %>%
  summarise(mean_num = mean(num, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = burnstatus, values_from = mean_num) %>%
  filter(!is.na(Burned) & !is.na(Unburned)) %>%
  mutate(diff_burn_nonburn = Burned - Unburned)

moisture_diff |>
  ggplot(aes(x = TIMESTAMP, y = diff_burn_nonburn, group = depth, color = depth)) +
  geom_line(size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_point(size = 1) +
  facet_grid(depth ~ .) +
  labs(x = "Date", y = "Moisture Difference (Burned - Nonburned)", title = "Moisture Diff (Burned vs Nonburned)") +
  theme_bw(base_size = 30)
```


```{r}
# Figure 8. Average moisture difference distributions by depth

moisture_diff |>
  ggplot(aes(x = diff_burn_nonburn)) +
  geom_density(fill = "purple", color = "purple") +
  facet_wrap(~ depth) +
  labs(
    x = "Moisture Difference (Burned - Nonburned)",
    y = "Frequency",
    title = "Moisture Difference Distribution (Burned vs Nonburned)"
  ) +
  theme_bw(base_size = 50)
```